Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
minima

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from xxx to string diagrams~\cite{bonchi2022string}.

@@

@@

The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).\todo{P: questa prima parte da riportare in parte in intro}

In this section we show that actually much of the theory can be retained: confluence of switching still holds, while globality of independence holds in a weak form, conceptually linked to the fact that fusions introduce a form of disjunctive causality. Finally, a canonical form for switching sequences can also be recovered.

We start by proving two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for showing properties (1) and (2) above.

>>>>

A noteworthy advantage is that DPO rewriting can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to devise 
the foundations of a rewriting theory that can be subsequently instantiated 
to the context of interest, thus avoiding the need of repeatedly proving
similar ad-hoc results in for each specific setting. In fact, the framework is
so general that it allowed to recast in the DPO approach many additional formalisms,
ranging from toposes~\cite{johnstone2007quasitoposes} to string diagrams~\cite{bonchi2022string}.

@@

DA RIUSARE\\
The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).

@@

%The characterisation of switch equivalence based on consistent permutations is the key to
%derive some simple but fundamental properties for linear rewriting systems, namely
%\begin{enumerate}
%\item \emph{confluence of switching}: when transforming derivation
%  sequences into equivalent ones by switching independent steps, the
%  result not depends on the order in which switchings are applied, but
%  only on the resulting permutation;
  %
%\item \emph{globality of independence}: if two independent steps are
%  moved forward or backward due to the application of switchings to a
%  derivation, they remain independent.
%\end{enumerate}
%
%
%In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).\todo{P: questa prima parte da riportare in parte in intro}
%
%The failure of the characterisation of switch equivalence via
%consistent permutations for left-linear rules leads to the
%question asking whether the properties above hold.

We argued in the introduction that the characterisation of switch equivalence based on consistent permutations 
is the key to derive some simple but fundamental properties for linear rewriting systems, namely
\emph{confluence of switching} and  \emph{globality of independence}.
In this section we show that actually much of the theory can be retained: confluence of switching still holds, 
while globality of independence holds in a weak form, conceptually linked to the fact that fusions introduce a form of disjunctive causality. 
Finally, a canonical form for switching sequences can also be recovered.

We start with two lemmas dealing with derivations of length
$3$. Despite being technical, these lemma are the fundamental building blocks for proving properties (1) and (2) above.
